\chapter{Background}
\label{sec:background}

% Background on design of DynamoRIO, all relevant considerations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DBI Frameworks}

In order to understand this thesis, it is important to understand what a dynamic
binary instrumentation (DBI) framework provides to a tool writer.

The primary job of a DBI framework is to interpret a native application as it
executes and provide an abstract representation of the program code that the
tool can analyze and instrument.  At first, it would seem easier to simply
disassmble the application in question and insert instrumentation code into it
statically.  However, for modern applications, this approach simply does not
work.  First, there are dynamically loaded libraries that the application may be
linked against.  Some of these are possible to identify, such as {\tt libc} or
others.  Some, however, may be dynamically loaded via other interfaces such as
{\tt dlopen} on Linux and {\tt LoadLibrary} on Windows.  These are not possible
to predict statically, and a static instrumentation tool will not be able to
observe and instrument these instructions.  Hence, a {\em dynamic} binary
instrumentation framework is needed to run alongside the application and
intercept every instruction that the application would have executed were it to
run natively.

Furthermore, a dynamic framework maintains control even in the face of such
convoluted techniques as self-modifying code.  As techniques such as embedded
Just In Time (JIT) compilation become more prevalent, it becomes more important
to be able to observe such dynamically generated code.  A DBI framework is also
responsible for providing all of the native operating system interfaces to the
application just as if the application were running natively.  This can be a
daunting challenge, as the operating system interface is large, and an
application can register many points of entry with the operating system such as
signal handlers.  A good DBI framework, such as DynamoRIO, Pin, or Valgrind,
will intercept all of these requests and ensure that control is maintained and
the tool author is able to observe all instructions.

Finally, a DBI framework provides transparency of the framework and the tool to
the application.  Even if the application uses introspection techniques such as
stack tracing or listing all the threads in the process, the application will
not notice the tool or the DBI framework running alongside it.

For all of these reasons, it is highly desireable to build dynamic program
analysis tools with DBI frameworks.  The goal of this thesis is to make it
easier to use DBI frameworks to write analysis tools that perform well.  For
this thesis, we chose to start by modifying DynamoRIO, which we describe in the
following section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DynamoRIO's Execution Model}

When DynamoRIO takes control, it sets up its own execution context, separate
from that of the application.  DynamoRIO's context consists of a separate stack,
and a thread-local data structure describing its state.  The application context
consists of the original application stack along with all of the registers that
DynamoRIO may clobber while executing its own code.  Once DynamoRIO has switched
to its own stack, it determines what the next application program counter would
have been and begins the process of interpretation.

DynamoRIO does not execute the original application instructions.  Instead, it
interprets the application code one basic block at a time.  For a previously
unencountered target application PC, it decodes instructions from the first
instruction until the next control transfer.  If DynamoRIO has loaded a program
analysis tool, it will present the instructions in the basic block to the tool
for instrumentation.  After instrumentation, DynamoRIO mangles the terminating
control transfer instruction to maintain control after the basic block finishes
executing.  Specifically, control flow instructions are mangled so that they
jump into DynamoRIO's central dispatch mechanism which will figure out what to
do next.

Once DynamoRIO and the analysis tool are done modifying the application
instruction stream, the instructions are encoded into a ``fragment'' in the
``code cache.'' The code cache is the memory space allocated by DynamoRIO for
translated application code.  This process is known as ``translation.'' Finally,
DynamoRIO switches back to the application context and starts executing the new
fragment.

When the basic block finishes execution, instead of transferring to the original
application target, it will re-enter the DynamoRIO VM with information about the
original target application program counter.  If the target application PC is
not in the code cache yet, DynamoRIO will then repeat the process of translation
for the next basic block.  After translation, it will return to the application
context and re-start execution from the freshly translated fragment.

Additionally, as perhaps the most important optimization in DynamoRIO, if the
control transfer target is direct, the two basic blocks will be ``linked''
together.  This is accomplished by modifying the terminating control transfer
instruction from the previous fragment to directly target to the beginning of
the new fragment.  Other bookkeeping structures are also updated to support
unlinking of fragments, which is important for supporting self-modifying code.
As a result, when a code path executes more than once, it will not have to leave
the application context to look up the fragment for the target PC in the
DynamoRIO VM.

% TODO: Perhaps some figures?

\section{Transparency}
\label{sec:transparency}

One of the primary requirements for DynamoRIO and systems like it is that the
application should run as if it were running on the native OS and hardware.  For
example, all control flow decisions made by the native application with the same
inputs should be exactly the same in the emulated execution.  Even a best effort
should be made to leave the memory layout completely undisturbed, or the native
execution will be different.  This means that DynamoRIO must also handle corner
cases like self-modifying code and stack crawlers as if they were running
natively.  DynamoRIO observes two guidelines to make this possible.

First, leave anything about the application unchanged if possible.  For example,
DynamoRIO never touches the application stack or even assumes that it is valid.
This complicates context switching out of the application and back to DynamoRIO,
which is described in Section \ref{sec:context_switching}.

Second, if it is not possible to leave the application unchanged, pretend that
it still is.  An example of this is that DynamoRIO needs to protect application
code from modification in order to maintain code cache consistency.  However,
the application must not know about these page protections, or it may change
behavior.  Therefore, DynamoRIO must intercept system calls that query this
state and emulate them as if the application were running natively.

There are many other transparency concerns that DynamoRIO abstracts away from
program analysis tool authors which are described at length in the previous
work.  Our main concern with transparency is the way that it constrains the
design of context switching, which is important for tool performance.
